# Description of run_analysis.R

The R script contained in this repository is used to process data obtained from
[this source](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones ) about human activity recognition using smartphones. 

In order to run run_analysis.R, download the UCI Har Dataset folder and the script and place them in the same directory. 

## Step 1: Read the Data into R. 

The following code will read all of the data sets into R to prepare for processing. 
```{r echo=TRUE}
xtest <- read.table("UCI Har Dataset/test/X_test.txt")
xtrain <- read.table("UCI Har Dataset/train/X_train.txt")
ytest <- read.table("UCI Har Dataset/test/Y_test.txt")
ytrain <- read.table("UCI Har Dataset/train/Y_train.txt")
subjectTest <- read.table("UCI Har Dataset/test/subject_test.txt")
subjectTrain <- read.table("UCI Har Dataset/train/subject_train.txt")
```
## Step 2: Merging the data sets. 
The first step I took is to combine the activities dataset. ytrain and ytest each have a single varible and many observations of it. 
```{r echo=TRUE}
activities <- rbind(ytrain, ytest)
```

In order to create a data set that contains descriptive names for the activities, this for loop will replace each number with the appropriate activity name. 

```{r echo=TRUE}
for (i in 1:nrow(activities)){
  if (activities[i,1]==1){activities[i,1] <- c("Walking")} 
  else if (activities[i,1]==2){activities[i,1] <- c("Walking_Upstairs")}
  else if (activities[i,1]==3){activities[i,1] <- c("Walking_Downstairs")}
  else if (activities[i,1]==4){activities[i,1] <- c("Sitting")}
  else if (activities[i,1]==5){activities[i,1] <- c("Standing")} 
  else{activities[i,1] <- c("Laying")}
}
```

Now I will merge the rest of the data sets together and combine them with the activities
data set. 

```{r echo=TRUE}
xy_test <- cbind(xtest, subjectTest)
xy_train <- cbind(xtrain, subjectTrain)
combined <- rbind(xy_train, xy_test)
data <- cbind(combined, activities)
```

Next we will add names to the columns from the features.txt file in the data

```{r echo=TRUE}
features <- read.table("UCI Har Dataset/features.txt")
features <- features[,2]
features_vector <- as.vector(features)
features_vector <- append(features_vector, c("subject", "activity"))
```

Now apply the names to the data frame.
```{r echo=TRUE}
names(data) <- features_vector
```

## Step 3: Subsetting the data to only keep columns referncing mean and standard deivation

Using grep, I will capture what columns to keep through partial matchng of column names. 
```{r echo=TRUE}
mean_names <- grep("mean()", colnames(data))
std_names <- grep("std()", colnames(data))
keep_cols <- c(mean_names, std_names, 562, 563)
```

Then we can subset the data
```{r echo=TRUE}
data_subset <- data[keep_cols]
```

## Step 4: Aggregating the data by subject and actviity
```{r echo=TRUE}
agg_subset <- aggregate(data_subset, list(data_subset$subject, data_subset$activity), FUN=mean)
tidy_data <- agg_subset[1:81]
```

The following code will clean up the clunky variable names by removing hyphens and parathensis. 

```{r}
neat_names <- c("tbodyacc_mean_x", "tbodyacc_mean_y", "tbodyacc_mean_z", "tgravityacc_mean_x","tgravityacc_mean_y", "tgravityacc_mean_z", "tbodyaccjerk_mean_x", "tbodyaccjerk_mean_y","tbodyaccjerk_mean_z", "tbodygyro_mean_x", "tbodygyro_mean_y", "tbodygyro_mean_z",
"tbodygyrojerk_mean_x", "tbodygyrojerk_mean_y", "tbodygyrojerk_mean_z", "tbodyaccmag_mean","tgravityaccmag_mean", "tbodyaccjerkmag_mean", "tbodygyromag_mean", "tbodygyrojerkmag_mean","fbodyacc_mean_x", "fbodyacc_mean_y", "fbodyacc_mean_z", "fbodyacc_meanfreq_x", "fbodyacc_meanfreq_y", "fbodyacc_meanfreq_z", "fbodyaccjerk_mean_x", "fbodyaccjerk_mean_y","fbodyaccjerk_mean_z", "fbodyaccjerk_meanfreq_x", "fbodyaccjerk_meanfreq_y", "fbodyaccjerk_meafreq_z","fbodygyro_mean_x", "fbodygyro_mean_y", "fbodygryo_mean_z", "fbodygyro_meanfreq_x","fbodygryo_meanfreq_y", "fbodygryo_meanfreq_z", "fbodyaccmag_mean", "fbodyaccmag_meanfreq","fbodyaccjerkmag_mean", "fbodyaccjerkmag_meanfreq", "fbodygyromag_mean", "fbodygryomag_meanfreq","fbodygyroherkmag_mean", "fbodygyrojerkmag_meanfreq", "tbodyacc_std_x", "tbodyacc_std_y","tbodyacc_std_z", "tgravityacc_std_x", "tgravityacc_std_y", "tgravityacc_std_z","tbodyaccjerk_std_x", "tbodyaccjerk_std_y", "tbodyaccjerk_std_z", "tbodygyro_std_x","tbodygyro_std_y", "tbodygyro_std_z", "tbodygyrojerk_std_x", "tbodygyrojerk_std_y","tbodygyrojerk_std_z", "tbodyaccmag_std", "tgravityaccmag_std", "tbodyaccjermag_stud","tbodygyromag_std", "tbodygyrojerkmag_std", "fbodyacc_std_x", "fbodyacc_std_y", "fbodyacc_std_z", "fbodyaccjerk_std_x", "fbodyaccjerk_std_y", "fbodyaccjerk_std_z", "fbodygyro_std_x", "fbodygyro_std_y", "fbodygyro_std_z", "fbodyaccmag_std",
"fbodyacjerkmag_std", "fbodygyromag_std", "fbodygyrojerkmag_std")
                
names(tidy_data) <- c("subject", "activity", neat_names)
```

Finally, we write the data to a file in the working directory called tidy_data.txt
```{r}
write.table(tidy_data, file="tidy_data.txt")
```